---
title: "Desctable"
output: github_document
---

```{r, echo = F, message = F, warning = F}
knitr::opts_chunk$set(message = F, warning = F)
```

[![Travis-CI Build Status](https://travis-ci.org/desctable/desctable.svg?branch=master)](https://travis-ci.org/desctable/desctable)  [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/desctable)](https://cran.r-project.org/package=desctable) [![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/desctable)](https://www.r-pkg.org:443/pkg/desctable)

**Warning to existing users**  
*This version introduces a new API that should make the creation of tables more flexible.  
The old API is still present but in a deprecated mode.  
See the roadmap below, and the website for the new usage.  
Suggestions about this change are welcome !*

---

# Introduction

Desctable aims to be a simple and expressive interface to building statistical tables in R.

# Installation

Install from CRAN with

```
install.packages("desctable")
```

or install the development version from github with

```
devtools::install_github("desctable/desctable")
```

# Basic usage

Load the package

```{r}
library(desctable)
```

Simply apply `desc_table` on a dataframe or a grouped dataframe to get a statistical table

```{r}
iris %>%
  desc_table()
```

Declare the statistics you want to see, and give them the name of your choice

```{r}
iris %>%
  desc_table("N" = length,
             "%" = percent,
             mean,
             sd)
```

Create comparative tables, compute statistical tests and output to `pander` for crisp markdown rendering!

```{r}
mtcars %>%
  dplyr::mutate(cyl = factor(cyl),
                vs = factor(vs, labels = c("V-shaped", "straight")),
                am = factor(am, labels = c("automatic", "manual"))) %>%
  group_by(am) %>%
  desc_table(N = length,
             "%" = percent,
             "Median" = median,
             IQR) %>%
  desc_tests(vs = ~chisq.test) %>%
  desc_output("pander")
```

Read more in the [vignette](articles/desctable.html) !
