---
title: "desctable tips"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{desctable tips}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F, message = F, warning = F}
library(desctable)
```

Here is collection of tips and tricks to go further with *desctable*

# 

## 

### Label variables

You can define labels for variables using the `.labels` argument in `desc_table`

```{r}
labels <- c(mpg   = "Miles/(US) gallon",
            cyl   = "Number of cylinders",
            disp  = "Displacement (cu.in.)",
            hp    = "Gross horsepower",
            drat  = "Rear axle ratio",
            wt    = "Weight (1000 lbs)",
            qsec  = "1/4 mile time",
            vs    = "Engine",
            am    = "Transmission",
            gear  = "Number of forward gears",
            CARBURATOR = "Number of carburetors")

mtcars %>%
  desc_table(.labels = labels) %>%
  desc_output("DT")
```

As you can see with `CARBURATOR` instead of `carb`, not all variables need to have a label, and unused labels are discarded.

### Default statistics

`desc_table` chooses its own statistics this way:

- always show `N = length`
- show `"%" = percent` if there is at least a factor
- show `min`, `max`, `Q1`, `Q3`, `median`, `mean`, `sd`, `IQR` if there is at least a numeric

### Defining your own default statistics

You can define your own automatic statistic function using the `.auto` argument in `desc_table`.  
This function should accept one argument, the table to choose statistics for (in the case of a grouped dataframe the subtables will be passed to the function). It should return a list of statistics.  
Here is the code of `stats_auto`, the default value of `.auto`

```{r, eval = F}
stats_auto <- function(data) {
  data %>%
    lapply(is.numeric) %>%
    unlist() %>%
    any -> numeric

  data %>%
    lapply(is.factor) %>%
    unlist() %>%
    any() -> fact

  stats <- list("Min"  = min,
                "Q1"   = ~quantile(., .25),
                "Med"  = stats::median,
                "Mean" = mean,
                "Q3"   = ~quantile(., .75),
                "Max"  = max,
                "sd"   = stats::sd,
                "IQR"  = IQR)

  if (fact & numeric)
    c(list("N" = length,
           "%" = percent),
      stats)
  else if (fact & !numeric)
    list("N" = length,
         "%" = percent)
  else if (!fact & numeric)
    stats
}
```

### Reuse a list of defined statistics

If you often reuse the same statistics for multiple tables and you don't want to repeat yourself, you can splice a list to `desc_table` using the `rlang::!!!` operator

```{r}
stats = list(N = length,
             Mean = mean,
             SD = sd)

mtcars %>%
  desc_table(!!!stats) %>%
  desc_output("DT")

```

When splicing, all stats need to be explicitly named

```{r}
stats2 = list(N = length,
              mean,
              sd)

mtcars %>%
  desc_table(!!!stats2) %>%
  desc_output("DT")

```

You can also define a "dumb" automatic function

```{r}
default_stats <- function(data)
{
  list(N = length,
       mean,
       sd)
}
```

### Default statistical tests

`desc_table` chooses its own statistical tests this way:

- if the variable is a factor, use `fisher.test`
  - if `fisher.test` fails, fallback on `chisq.test`
- if the variable is numeric, use
  - `wilcoxon.test` if there are two groups
  - `kruskal.test` if there are more than two groups

### Defining your own default statistical tests

You can define your own automatic statistic function using the `.auto` argument in `desc_tests`.  
This function should accept two arguments, the variable to compare and the grouping variable, and return a statistical test that accepts a `formula` argument and returns an object with a `p.value` element.  
Here is the code of `tests_auto`, the default value of `.auto`

```{r, eval = F}
tests_auto <- function(var, grp) {
  grp <- factor(grp)

  if (nlevels(grp) < 2)
    ~no.test(.)
  else if (is.factor(var)) {
    if (tryCatch(is.numeric(fisher.test(var ~ grp)$p.value), error = function(e) F))
      ~fisher.test(.)
    else
      ~chisq.test(.)
  } else if (nlevels(grp) == 2)
    ~wilcox.test(.)
  else
    ~kruskal.test(.)
}
```

You can also provide a default statistical test using the `.default` argument

```{r}
mtcars %>%
  group_by(am) %>%
  desc_table(mean, sd) %>%
  desc_tests(.default = t.test) %>%
  desc_output("DT")
```

You can still choose individual tests when you define either a `.auto` or a `.default` test

```{r, warning = F}
mtcars %>%
  group_by(am) %>%
  desc_table(mean, sd, median, IQR) %>%
  desc_tests(.default = t.test, carb = wilcox.test) %>%
  desc_output("DT")
```

Note that if a `.default` test is provided, `.auto` is ignored.

### Output options

You can set the number of significant digits to display with the `digits` argument.
The p values are truncated at 1E-digits.

```{r}
iris %>%
  group_by(Species) %>%
  desc_table(mean, sd) %>%
  desc_tests() %>%
  desc_output("DT", digits = 10)
```

Any additional argument given to `desc_output` will be carried to the output function

```{r}
iris %>%
  group_by(Species) %>%
  desc_table(mean, sd) %>%
  desc_output("DT", filter = "top")
```
